Paper 1: Triangulating trends in inequality around the world from secondary databases

### 2023-04-26 PhD_pages/Paper_1/paper1.qmd
Paper 1: drafting

Two sources of data have been particularly impactful in this regard. One is the World Bank's harmonized collection of household survey data, published online through its [*Poverty and Inequality Platform (PIP)*](https://pip.worldbank.org/). Another is the [*World Inequality Database (WID)*](https://wid.world/), which emerged out of the burgeoning literature on top incomes th incorporation of tax data [@atkinson_top_2014].

These two sources are the only regularly-updated databases providing global coverage of distributional data on a relatively comparable basis.^[TODO: Both datasets suffer important limitations with respect to comparability acrosss time and location. These will be discussed in the data section below. Both datasets, however, take significant steps to harmonize, *ex-post*, the available underlying data. In this respect they can be distinguished from databases such as WIID and SWIID [@atkinson_promise_2001]]

Both have played substantial agenda-setting roles in academic research, policy and public debate. Their use and impact, however, has been confined to largely separate arenas, reflecting differences in the concepts of inequality they employ and the coverage they provide.

The *PIP* data has typically been the favoured choice for investigating the whole distribution of welfare across households, particularly at a global scale. () The *WID* data at the top, focus on role of capital incomes.

Moreover, the many methodological differences separating the two datasets prevent users combining the insights that they each offer individually into a coherent empirical view informed by both. In choosing datasets, we are simultaneously choosing underlying data source, methods, welfare concept, and coverage. The two datasets represent distinct and, to a large extent, incommensurable 'worlds' of inequality measurement.

This disconnect is all the more pertinent because they disagree on basics. Firstly. Secondly.

Very confusing terrain for people to understand. We don't know which of the two worlds we live in. Depends on a choice about which little can be determined.

This paper is the first of two which will try to ocnnect them up.

Section 2. Section 3...

### Two 'worlds' of inequality measurement

>>>>

Two sources of data have been particularly impactful in this regard. One is the World Bank's harmonized collection of household survey data, published online through its [*Poverty and Inequality Platform (PIP)*](https://pip.worldbank.org/). Another is the [*World Inequality Database (WID)*](https://wid.world/), which emerged out of the burgeoning literature on top incomes [@atkinson_top_2014].

These two sources are the only regularly-updated databases providing global coverage of distributional data on a relatively comparable basis.^[TODO: Both datasets suffer important limitations with respect to comparability acrosss time and location. These will be discussed in the data section below. Both datasets, however, take significant steps to harmonize, *ex-post*, the available underlying data. In this respect they can be distinguished from databases such as WIID and SWIID [@atkinson_promise_2001]] Both have played substantial agenda-setting roles in academic research, policy and public debate.

Their use and impact, however, has taken place within largely separate arenas. Many methodological differences separate the two datasets, including the  underlying data sources they rely on, the methods and assumptions applied to those sources, the welfare concept, the summary indices of inequality favoured, and coverage they provide. This lack of alignment between the datasets greatly limits our ability to combine the insights provided by each into a coherent empirical view informed by both. The two datasets represent distinct and, to a large extent, incommensurable 'worlds' of inequality measurement.

Moreover, these two worlds of measurement yield different views concerning the nature and evolution of inequality.

Firstly, while the *WID* data and its associated global reports suggest rising inequality was a near-ubiquitous phenomenon [@alvaredo_world_2018; @chancel_world_2021], in contrast the World Bank *PIP* data and its associated reports offer a considerably more mixed picture, in which inequality has been rising and falling in roughly equal numbers of countries [@world_bank_poverty_2016].[^WID_PIP_quotes]

[^WID_PIP_quotes]:
    > Income and wealth inequalities have been on the rise nearly everywhere since the 1980s, following a series of deregulation and liberalization programs which took different forms in different countries. [@chancel_world_2021, pp. xx]

    > [Since c.1993] the number of countries with rising inequality was **slightly higher** than the number of countries with falling inequality. Yet, the average Gini among these 91 countries actually **declined**. [@world_bank_poverty_2016, pp. xx]

Secondly, the datasets provide contrasting accounts as to how inequalities *within* countries compare to the differences in average incomes seen *between* countries. Most analyses of global income inequality to date have relied heavily on the *PIP* data [@Kanbur2022-lk; @lakner_global_2016; @anand_global_2015; @milanovic_global_2015; @Nino-Zarazua2017-tv; @Ravallion2014-mk]. A consistent result from these studies is that, despite recent convergence [@patel_new_2021], differences in average incomes between countries continue to make up the lion's share of global inequality. On such accounts, a person's income is largely determined by the country in which they happen to have been born [@milanovic_global_2015]. More recent analysis of the *WID* data, however, flips this established result on its head. @chancel_global_2021 find within-country inequalities instead 'dominate' inequality between countries, contributing 78% in a decompositon of the global Theil index.

Thus despite the greatly improved range of distributional data at our disposal, considerable ambiguity continues to surround some of the most basic descriptive questions concerning inequality around the world.

This is the first of a pair of papers that try to address this problem by bridging between the *PIP* and *WID* datasets, and the two 'worlds' of measurement they manifest. The novelty of the papers lies preciesly in considering both datasets in conjunction, in contrast to the many studies that consider only one in isolation. In taking such a combined view, we are able to 'triangulate' a more robust picture of inequality around the world by highlighting where the datasets agree and reinforce one another, and pinpointing more precisely where and why they do not.

The present paper the focuses on trends in within-country inequality. A companion paper considers overall global inequality, factoring in differences in average incomes between countries alongside the within-country inequalities analysed here.

The rest of the paper is organized as follows. Section 2 provides a more detailed description of the *PIP* and *WID* datasets and their main differences. Section 3 compares within-country inequality trends in the two datasets, starting with the metrics where they fall furthest apart. For this we use the Gini of disposabale income (or consumption) and the top 1% share of pretax income as metrics emblematic of the *PIP* and *WID* datasets respectively.
Section 4 takes a first simple step to bridge betweent he datasets, by aligning the coverage. Section 5 considers a range of inequality indices, comparing the datasets on a like-for-like basis.
Section 5 zooms in subset of countries where there is most disagreement. the role of welfare metrics and specific data quality issues for 
Section 6 concludes, drawing out the main overlaps and disagreements between the two datasets, accounting for the steps to

# Two 'worlds' of inequality measurement
